\documentclass[10pt, conference, compsocconf]{IEEEtran}

\usepackage{color}
\usepackage{soul}

%\documentclass[a4paper, UKenglish]{lipics}
%\usepackage{amsmath}
%\theoremstyle{plain}

\usepackage{hyperref}
\usepackage[noend]{algorithm2e}
\usepackage{algorithmicx}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{times}
\let\proof\relax % undefine the environment
\let\endproof\relax
\usepackage{amsthm}

\newcommand{\todo}[1]{\color{red}\textbf{\hl{\{ #1 \}}}\color{black}\xspace}
\newcommand{\ceil}[1]{\left\lceil #1 \right\rceil}

\begin{document}
\title{High Performance Space-Time Kernel Density Estimation}


\author{\IEEEauthorblockN{Erik  Saule$^\dag$, Dinesh Panchananam$^\dag$, Alexander Hohl$^\ddag$, Wenwu Tang$^\ddag$, Eric Delmelle$^\ddag$}
\IEEEauthorblockA{$^\dag$ Dept. of Computer Science, $^\ddag$Dept. of Geography and Earth Sciences\\
UNC Charlotte\\
Charlotte, NC, USA\\
Email: {esaule,dpanchan,ahohl,wtang4,eric.delmelle}@uncc.edu}
}

\maketitle


\begin{abstract}
  The exponential growth of available data, caused by the BigData era,
  increased the interest in interactive exploratory analysis as dataset
  can no longer be understood by crawling manually through it and
  looking at a few statistics. In a Geographical Information Systems
  (GIS), the dataset is often composed of events localized in space
  and time; and visualizing such a dataset involves building a map of
  where the events occurred.

  We focus in this paper on events that are localized among three
  dimensions (latitude, longitude, and time), and on computing the
  first step of the visualization pipeline, space-time kernel density
  estimation, which is most computationally expensive.  Starting from
  a gold standard implementation, we show how algorithm design and
  engineering, parallel decomposition, and acceleration can be applied
  to bring real-time computing to space-time kernel density
  estimation. We validate our techniques on real world datasets
  extracted from infectious disease, social media, and ornithology.
\end{abstract}

\begin{IEEEkeywords}
space-time kernel density; performance; distributed memory; gpu; shared-memory; scheduling
\end{IEEEkeywords}


\IEEEpeerreviewmaketitle

\section{Introduction}

\section{Space-Time Kernel Density Estimation}

\subsection{Motivation}

\todo{insert here some text to describe stkde, why it is important,
  and where it applies.}

\subsection{Problem Definition}


Notations: domainsize $gx$, $gy$, $gt$. bandwidth: $sbw$,
$tbw$. Spatial resolution $sres$, temporal resolution $tres$. This
forms a grid of size $GX=\ceil{\frac{gx}{sres}}$,
$GY=\ceil{\frac{gy}{sres}}$, $GT=\ceil{\frac{gt}{tres}}$. And we
denote the bandwidth in voxel $SBW=\ceil{\frac{sbw}{sres}}$,
$TBW=\ceil{\frac{tbw}{tres}}$.

There are $N$ observations.

\subsection{Gold Standard Implementation}

The gold standard implementation follows the exact definition of stkde
as it is given above. It is a voxel-based algorithm we call VB. For
each voxel, VB finds the close observations and calculate contribution
to the density of this particular voxel.

\todo{include voxel based algorithm}

The algorithm does $\theta(GX * GY * GT * N)$ distance test and computes
$\theta(N * SBW^2 * TBW )$ density. Noticing that the bandwidth has to be
smaller than the grid size, the complexity of the algorithm is
$\theta(GX * GY * GT * N )$.

Requires $\theta(GX * GY * GT)$ memory.

\subsection{Related Works}

\section{Algorithm Design and Engineering}

\subsection{Observation-based Algorithm}

The voxel-based algorithm suffers from a massive cost of computing
distances. The gold implementation reduces them by extremely
decomposition the domain, but that still incures a significant
overhead. These calculation are mostly unnecessary since we know
exactly where each observation will radiate density. The observation
based algorithm OB leverages that property.

\todo{spell out algorithm}

The algorithm OB incurs mostly two costs. Initializing the memory which costs
$\Theta(GX * GY * GT)$ and computing the impact of each observation
$\Theta(N * SBW^2 * TBW)$. Note that it is possible that either of these
two term is significantly larger than the other. Therefore the
complexity of the algorithm is $\Theta(GX * GY * GT + N * SBW^2 * TBW)$.

\subsection{Exploiting Symmetries}

When the bandwidth is large computing the density contribution of each
observation each voxel in the bandwidth is the most expensive
operation. Each of these $N * (2 * SBW)^2 * 2 * TBW$ calculations costs
approximately 40 floating point operations. (The exact cost is
difficult to estimate since there are floating point additions,
multiplications, divisions, and square root operations.)

One can remark that the calculation is mostly redundant. \todo{give
  precise formula.}  Indeed for one particular observation, its
contribution to the neighbooring vertices can be decomposed in two
components. One that only depends on the spatial coordinate of the
voxel with respect to the observation and does not depend on its
temporal coordinate. And one that only depends on the temporal
coordinate and not on the spatial coordinate.

\todo{include figure.}

Therefore, one can write an algorithm to leverage these symmetries in
the problem. We wrote three variants, the OB-DISK variant ensures that
the invariant on the spatial domain is only computed once. The OB-BAR
variant ensures that the invariant on the temporal domain is only
computed once. The SYM variant computes independently the spatial
invariant and the temporal invariant and then adds the density caused
by the observation by taking the product of the two invariants. Notice
that leveraging the symmetries in the calculation does not change the
complexity. The SYM algorithm still has a complexity of $\Theta(GX *
GY * GT + N * SBW^2 * TBW)$. However the number of flops reduced to 
\todo{estimate flop cost.}

\todo{include pseudo code of the SYM variant.}

Notice that this exploitation of symmetries is not possible in a voxel
based algorithm.

\section{Domain-based parallelism}


\subsection{Domain Replication}

The simplest way to parallelize this kind of computation is to split
the observation equally among the $P$ different computational
units. Though if two close-by observations are computed
simultaneously, the cylinder of density around it might collide and
there is a race condition.

The SYM-DR algorithm solves the data race issue by having
each of the $P$ computational unit aggregate the result on a local
copy of the domain. Once all the observations have been processed, the
$P$ copies need to be summed.

This algorithm has a memory requirement of $\Theta (P * GX * GY * GT)$
and a parallel amount of work of $\Theta (P * GX * GY * GT + N * SBW^2
* TBW)$. Though the increase in work enables the parallelism to be
pleasingly parallel. (It is actually in two pleasingly parallel phases,
processing the observations and reducing the partial results.)

\todo{give pseudo code of the algorithm}

\subsection{Domain Decomposition}

An other strategy to avoid the data race is by computing the intensity
of the voxels of different subdomains independently. The SYM-DD
algorithm splits the domain in AxBxC subdomains. And each subdomain is
associated the observations which cylinder intersects with a voxel of
the subdomain. Then the algorithm proceeds with processing each
subdomain independently by only considering the impact of the
observation on the voxels within the subdomain.

\todo{include algorithm}

The main problem with Domain Decomposition is that it requires some
observations to be replicated across multiple subdomain which incurs
additional work. Indeed for a particular observation replicated in two
subdomains (if it is replicated in more, a similar effect happens),
its cylinder is split across two subdomain. Assume the split is
temporal, then using SYM each subdomain will need to compute part of
the temporal invariant, the entire spatial invariant, and part of the
cylinder. Therefore both subdomain needed to compute the entire
spatial invariant.

\todo{draw figure that examplify the issue}

If the algorithms keeps the number of subdomain small such that each
observation is not replicated more than a constant number of time then
the work performed when expressed in Big-Oh notation does not change.
(This can be achieved, for instance, by having subdomains larger than
$SBW \times SBW \times TBW$.) However, the practical amount of work can
increase by a non negligable factor. (We will see factors of about 1.5
in the experiments.)

An other issue with Domain Decomposition is load imbalance. The
observations are unlikely well distributed in the domain space, but
more likely clustered around some locations. The subdomain containing
such a location will have a significantly higher work than other
subdomains. And this work is not executed in parallel. One could
decompose the domain more, but replication observation (described
above) will incur an even higher overhead which may end up nullifying
the gain in load balance.

\section{Observation-based parallelism}

\subsection{Observation Decomposition}

Both of the previous strategies increase the amount of work
significantly. We want to achieve parallelism without cutting a
cylinder and without requiring each thread to have its own memory
space. To achieve that goal, we propose the SYM-OD algorithm that
decomposes the observations in sets that can be safely performed
simultaneously.

As long as two observations are separated by $2*sbw$ in a spatial
dimension or by $2*tbw$ in the temporal dimension then the cylinder
induced by the two observations will not overlap. SYM-OD decomposes
the observations in $A \times B \times C$ subdomain of identical
size. As long as these subdomains are larger than $2*sbw$ in a spatial
dimension and $2*tbw$ in the temporal dimension, then the observations
in domain $x, y, z$ can be done concurently with the observations in
$x+2, y, z$.

\todo{include figure that show the decomposition}

We implemented a first version of this algorithm that organizes the
subdomain in 8 sets where the first set is composed of the subdomains
$2*i, 2*j, 2*k$, the second of all the subdomains $2*i+1, 2*j, 2*k$,
the third of all the subdomains $2*i, 2*j+1, 2*k$, etc.. The algorithm
processes the sets the one after the other by doing each subdomain
within a set in parallel. In practice, this is implementation using
$8$ OpenMP parallel for construct.

\todo{include algorithms}


\subsection{Coloring and Scheduling}

Note that such an implementation overconstrains the execution of the
algorithm. Indeed, despite they are not in the same set, subdomain
$1,0,0$ and $64, 64, 64$ can be computed at the same time. The real
constraint is that the observations contained in a subdomain can be
processed safely as long as none of the observation contained in a
neighboring subdomain is being processed. Using the OpenMP tasking
construct with dependency introduced in OpenMP 4.0 \todo{cite
  openmpspec}, one can precisely express this constraint, which is a
27-point stencil constraint.

Formally speaking, this problem is a combination of multiple very
classical graphs and scheduling problem. Modeling the subdomains as a
27-point stencil graph, identifying sets of subdomains to perform in
parallel is coloring the vertices of the graph so that no two
neighboring vertices share the same color. It differs from classical
graph coloring problem (see survey in~\cite{}) in that the objective
is not to minimize the number of color or to balance the number of
vertices of each color but to minimize the execution time of the
implied schedule.



one can extract, out of a
coloring of the subdomains, a dependency graph of which subdomain.

Coloring and moldable scheduling.



\section{Experiments}

\subsection{Experimental Setting}

\paragraph{Execution environment} We are using the copperhead machine
at UNC Charlotte to perform our experiments. A node of Copperhead is
composed of two Intel Xeon E5-2667 v3 processors. There is a total of
16 cores clocked at 3.2Ghz. Despite each core is capable of hosting 2
hyperthreads, hyperthreading was disabled. Each node is equipped with
128GB of \todo{DDR4 clocked at something Ghz}. The node uses RHEL 7.2
and Linux 3.10. All codes are compiled using gcc 5.3, which implements
OpenMP 4.0, with optimization flags \texttt{-O3 -march=native -fopenmp
  -DNDEBUG}. All reported execution time exclude all IOs.

\paragraph{Dataset} Three dataset are used in our experiments. Their
key attributes are summarized in table~\ref{tab:instance}. The Dengue fever
dataset~\cite{delmelle2013spatio}\todo{more}.

The Pollen dataset is composed of 588K tweets of US users from
February 2016 to April 2016 related to Pollen (that mention keywords
such as ``pollen'' and ``allergy''). Tweets that did not contain a
precise localization were approximated by picking a random location in
the approximated region provided by
Gnip\footnote{\url{http://www.gnip.com/}}. Tweets are mostly located
in the US, but there are tweets all over the northern hemisphere in
this dataset. Therefore, this is the region we modeled with a spatial
resolution ranging from $.2^\circ$ to $1^\circ$ and a temporal
resolution of 1 day.

The Flu dataset was obtained from the Animal Surveillance
database of the Influenza Research
Database\footnote{\url{https://www.fludb.org/}} on October 3rd,
2016. The dataset contains observations of birds tested positive for
carrying any strain the avian flu from 2001 to 2016. The birds were
observed all over the world and the modeled region spans as West as
Alaska and East as Japan, as South as the southern shore of Australia
and as North as the Northern shore of Russia. The domain is modeled
with spatial resolution from $.2^\circ$ to $1^\circ$ and a temporal
resolution of 1 day.

\begin{table}
  \input{analysis/instance_table.tex}
  \caption{Properties of the datasets. The domain is expressed in voxels
    and in MB. Spatial and temporal bandwidth are given in voxels.}
  \label{tab:instance}
\end{table}


\subsection{Algorithm design}

\begin{table*}
  \begin{tabular}{|l|rrrr|r|}
    \input{analysis/seq_table.tex}
  \end{tabular}
  \caption{Runtime of different algorithm. Times are in seconds. PB-SYM is the fastest.}
\end{table*}


\subsection{Domain-based parallelism}

\begin{table*}
  \begin{tabular}{|l|r|rrrrrrrrr|}
    \input{analysis/parspace-t1.tex}
  \end{tabular}
  \caption{Different spatial decomposition at t=1. Times are in
    seconds. Some decomposition improve performance dues to better
    cache fitting. Overdecomposition can cause significant overhead.}
\end{table*}

\begin{table*}
  \begin{tabular}{|l|r|rrrrrrrrr|}
    \input{analysis/parspace-t16.tex}
  \end{tabular}
  \caption{different spatial decomposition at t=16}
\end{table*}

\begin{table*}
  \begin{tabular}{|l|r|rrrrr|}
    \input{analysis/parpoint_table.tex}
  \end{tabular}
  \caption{domain replication. Times are in seconds. Missing data points run out of memory.}
\end{table*}

\subsection{Observation-based parallelism}

\begin{table*}
  \begin{tabular}{|l|r|rrrrrrr|}
    \input{analysis/parcol-t16.tex}
  \end{tabular}
  \caption{obs decomposition}
\end{table*}

\begin{table*}
  \begin{tabular}{|l|r|rrrrrrr|}
    \input{analysis/obssched-t16.tex}
  \end{tabular}
  \caption{obs decomposition - sched}
\end{table*}

\begin{table*}
  \begin{tabular}{|l|r|rrrrrrr|}
    \input{analysis/obscolorsched-t16.tex}
  \end{tabular}
  \caption{obs decomposition - color sched}
\end{table*}

\begin{table*}
  \begin{tabular}{|l|r|rrrrrrr|}
    \input{analysis/par_obs_mold-t16.tex}
  \end{tabular}
  \caption{obs decomposition - color sched mold. The missing item run out of memory becasue of a too high replication factor.}
\end{table*}



\section{Conclusion}

\section*{Acknowledgment}





\bibliographystyle{IEEEtran}
\bibliography{ipdps}


\end{document}


